### 2004-04-21 Documentation/part2.tex
part of the edm documentation

\section{Source properties}

@@

\section{Detector properties}

@@

Most advanced step: How to create a new experiment from scratch.

>>>>

\chapter{General concept}
\section{Source properties}

\GLOBES\ handles only point sources, thus it is not possible
to study effects from the finit size of the neutrino production
region like in the Sun. Furthermore only one source for each experiment
is possible, this make it impossible\footnote{at least rather tricky} 
to simulate multiple source experiments like KamLAND. All neutrino
sources useable within \GLOBES\ therefore can be specified by the
flux spectra for each flavour and CP sign and the total luminosity
of the source. There are two principal ways to provide a flux:
use a built-in source or a file provided by the user.

The built-in source is muon decay, since the neutrino
spectra and flavour composition are knon analytically in that case. For 
this sources the energy of the parent particle has to be specified
and the number of useful decays per year.

In the latter case the fluxes are read from a file with seven rows and
501 lines.
\begin{quotation}
$ E\quad
\Phi_{\nu_e}\quad
\Phi_{\nu_\mu}\quad
\Phi_{\nu_\tau}\quad
\Phi_{\bar\nu_e}\quad
\Phi_{\bar\nu_\mu}\quad
\Phi_{\bar\nu_\tau}$
\end{quotation}
where the energy values are evenly spaced. 
For accessing fluxes at 
arbitrary energies linear interpolation is used. Once the energy leaves the
range of values given in the file $0.0$ is returned.

Since many sources have two modes of operation, it is possible to define
a pair of fluxes by using the suffixes ``plus'' and ``minus'' in the file 
names. The individual members of the pair can be accessed by the polarity
variable. Also the built-in fluxes provide this choice.

For sources based on $\pi$-decay and nuclear reactors the total power
involved defines how neutrino are produced therefore this variable is
used to define the luminosity.

Furthemore it is useful to define the operation time of an experiment
in years.

@@

\subsection{Baseline \& matter profile}
The purpose of this module is to specify the path the neutrinos
are traveling. The central piece of information is the total length
of the path - the baseline\index{Baseline}.
\begin{equation}
L\,\left[\mathrm{km}\right]
\end{equation}

Furthermore there is
the matter profile encountered along the baseline which can be set in
several ways. The simplest way is to use a constant density corresponding
to the average density along the baseline calculated using the 
PREM\cite{Stacey} onion shell model of the 
Earth\index{PREM}\index{Earth matterdensity}.

Another possibility is to use several layers of constant density each in 
order to improve the approximation to the actual Earth matter profile. In this
case the user specifies the total baseline and the number of layers. The
time consumed in computing the oscillation probabilities is directly 
proportional to number of layers.

The third possibility to specify the matter profile is to provide a list
of thicknesses and densities for the layers.

For any type of profile a central value $a$ and a corresponding error 
$\delta a$ have to be specified\index{Density error}\index{Density center}.
\begin{equation}
\label{eq:density_error}
\rho(x)=a\cdot\rho_0(x)\,\left[\frac{\mathrm{g}}{\mathrm{cm}^2}\right]\,
\end{equation}
where $\rho_0(x)$ is the matter profile as chosen by the user and $\rho(x)$ is
the one used in the calculation of the oscillation probabilities. The default
value for $a$ is $1$ and a typical value for $\delta a = 0.05$.

\section{Detection properties}

The most obvious property of the neutrino detection is the total
active mass of the target.

\subsection{Cross section}
\label{sec:cross_section}

Cross sections\index{Cross sections} are used as part of the 
channel definition 
(see section~\ref{sec:channel}) and have to be provided by the user in form
of a file. Cross section are in general given as differential cross divided
by the energy
\begin{equation}
\hat\sigma(E)=\sigma(E)/E\,\left[ 10^{-38}\,
\frac{\mathrm{cm}^2}{\mathrm{GeV}^2} \right]
\end{equation}
The cross section files have seven rows and $1001$ lines.
\begin{quotation}
$log_{10} E\quad
\hat\sigma_{\nu_e}\quad
\hat\sigma_{\nu_\mu}\quad
\hat\sigma_{\nu_\tau}\quad
\hat\sigma_{\bar\nu_e}\quad
\hat\sigma_{\bar\nu_\mu}\quad
\hat\sigma_{\bar\nu_\tau}$
\end{quotation}
where the logarithm of the energy values is evenly spaced. 
For accessing cross section at 
arbitrary energies linear interpolation is used. Once the energy leaves the
range of values given in the file $0.0$ is returned.

\subsection{Channels}
\label{sec:channel}

Channels\index{Channel} represent an intermediate level in bewteen 
the pure oscillation 
physics given by the oscillation probablity $P_{\alpha\beta}$ and the
detected signal and background. A channel basicaly decsribes the path from one
initial state in the source to one interaction type (IT) in  the detector.  
Therefore a channel consists of the description of the inital state given 
by the polarity 
of the source, \eg\ muons or anti-muons in a neutrino factory, by the CP 
eigenvalue of the intial state, \ie\ neutrino or anti-neutrino and by the 
intial flavour. A channel furthermore consists of the interaction type in the
detector given by the final flavour of the neutrino, the interaction cross
section and the energy resolution function.

For the calculation of event rates, the first step is to compute the number of
events for each IT in the fiducial mass of the detector for each 
neutrino flavor and energy bin. The second step is to include the detector 
effects coming from
the insufficient knowledge used in the event reconstruction. We combine these
two steps in order to obtain the differential event rate spectrum for each
flavor and IT as seen by the detector, which we call a ``channel''. 
The channels
for all ITs then have to be combined in a way which takes into account
that the ITs or flavors can, either because of
physical reasons (\eg, the flavor--blindness of NC interactions) or 
because of detector effects (\eg, charge misidentification), not be 
 separated cleanly.

The master formula for the differential event rate for each channel 
is given by
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{eqnarray}
\label{eq:master_event}
\frac{dn_{\beta}^{\text{IT}}}{dE'}=&&N\,\int \int dE\,d\hat{E}\quad
\underbrace{\Phi_{\alpha} (E)}_{\mathrm{Production}} \times \nonumber\\
&&\underbrace{\frac{1}{L^2} P_{(\alpha\rightarrow\beta)}(E,L,\rho;\theta_{23},
\theta_{12},\theta_{13},
\Delta m^2_{31},\Delta m^2_{21},\deltacp)}_{\mathrm{Propagation}}
\times \nonumber \\ &&\underbrace{\sigma^{\text{IT}}_f(E)
k_f^{\text{IT}}(E-\hat{E})}_{\mathrm{Interaction}} \times \nonumber \\
&&\underbrace{ T_f(\hat{E}) V_f(\hat{E}-E')}_{\mathrm{Detection}}\,,
\end{eqnarray}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
where $\alpha$ is the initial flavor of the neutrino and 
$\beta$ is the final flavour, 
$E$ is the incident neutrino energy, $\Phi_{\alpha} (E)$ is the flux of the 
initial flavor at the
source, $L$ is the baseline length, $N$ is a normalization factor, and 
$\rho$ is the matter density. The interaction term is composed of 
two factors, which are the total cross section 
$\sigma^{\text{IT}}_\beta(E)$ for the flavor $f$ and
the interaction type IT, and the energy distribution of the 
secondary particle $k_\beta^{\text{IT}}(E-\hat{E})$ with 
$\hat{E}$ the energy of the secondary particle. The detector properties are 
modeled by the threshold function $T_\beta(\hat{E})$, coming from the the 
limited resolution or the cuts in the analysis, and the energy resolution 
function $V_\beta(\hat{E}-E')$ of the secondary particle. Thus, $E'$ is the 
{\em reconstructed} neutrino energy.

Since it is rather cumbersome to numerically solve this double integral,
we split the two integration. We first evaluate the integral over
$\hat{E}$, where the only terms containing $\hat{E}$ are
$k_\beta^{\text{IT}}(E-\hat{E})$,  $ T_\beta(\hat{E})$, and 
$ V_\beta(\hat{E}-E')$, and define:
\begin{eqnarray}
\label{eq:e_res} 
R_\beta^{\text{IT}}(E,E')\,\epsilon_\beta^{\text{IT}}(E')
 \equiv
\int d\hat{E} \quad T_\beta(\hat{E})\,k_\beta^{\text{IT}}(E-\hat{E})
\,V_\beta(\hat{E}-E')\,, 
\end{eqnarray}
Thus $R_\beta^{\text{IT}}(E,E')$ described the energy response of 
the detector, \ie\ a neutrino with a (true) energy of $E$ has a probablity
of $R_\beta^{\text{IT}}(E,E') dE'$ to have a reconstucted energy 
in bewteen $E'$ and $E'+dE'$. The function $R(E,E')$ is also refered to
as energy resolution function or smearing matrix in its actual representation
in the program\index{Energy resolution}\index{Smear matrix}. The detailed
defintion and initialization of the energy resolution function is described
in section~\ref{sec:energy}.

\subsection{Rules}
\label{sec:rules}

The set of rules\index{Rule} for an experiment is the final 
link between the event rate
computation and the statistical analysis. The information given by the rules
specifies how the $\chi^2$ is computed based on the raw event rates as 
described by the channels including possible systematical errors. 
Therefore a rule has two parts: the first part describes how signal and 
background events are composed out of the channels and the second part
specifies which systematical errors are considered and what the the size
of the relevant errors is.

Based on the definition of channels it is possible to assemble the 
signal and background event numbers for each bin.  
The number of events in one channel and  in one bin\index{Bin} $i$ is given by
\begin{equation}
n_i^c=\int_{E_i-\Delta E/2}^{E_i+\Delta E/2} dE' \quad
\frac{dn_{\beta}^{\text{IT}}}{dE'} (E')
\end{equation}
where $c$ is the channel index.
Note that the events are binned according to their \emph{reconstructed} energy.
The signal $s_i$ in one bin for a rule now can be composed out of 
several channels by
\begin{equation}
s_i=\epsilon_{c_1}\cdot n_i^{c_1}\,+\,\epsilon_{c_2}\cdot n_i^{c_2}\,+\,\ldots
\end{equation}
where the $\epsilon$'s are weight factors determined by the properties
of the detector.
Simlarly the background in one bin $i$ can be specified by
\begin{equation}
b_i=\epsilon_{c_1}\cdot n_i^{c_1}\,+\,\epsilon_{c_2}\cdot n_i^{c_2}\,+\,\ldots
\end{equation}

The set of all $s_i$'s and $b_i$'s now can be composed in a form that takes
some types of systematical error into account. The two most important and
most easily parameterized systematics are a normalization and an energy
calibration error. These error are considered seperately for the signal events
and the background events. The implementation of the normalization error
is straight forward
\begin{equation}
s_i(a):=a\cdot s_i
\end{equation} 
with an anloguous definition for the background events.

For the parameterization of an energy calibration error two possibilities
are implemented. The first one being somewhat simpler and second one
being more accurate. The first option is
\begin{equation}
s_i(a,b):=s_i(a)+b\cdot s_i\, E_i/(E_\mathrm{max}-E_\mathrm{min})
\end{equation}
and it is refered to as tilt since it describes a linear distortion, a tilt
of the event rate spectrum. The second option is closer to an actual energy
calibration error and is an approximation to the exact form
\begin{equation}
s_i(b) = \int_{(1+b)(E_i-\Delta E/2)}^{(1+b)(E_i+\Delta E/2)} dE' \quad
\frac{dn_{\beta}^{\text{IT}}}{dE'} ( E')
\end{equation}
or something like this, to be continued ...

Thus the total event rate $x_i$ in a bin $i$ is given by
\begin{equation}
x_i(a,b,c,d)=s_i(a,b)+b_i(c,d)
\end{equation}
and thus a function of four parameters. The central values for all of
the four parameters are always needed. They are called signal normalization
($a$), signal tilt/calibration ($b$), background  normalization ($c$) and
background tilt/calibration ($d$). The default values are
\begin{equation}
a=1\,,\quad b=0\,,\quad c={\tt NaN}\,,\quad d=0\,,
\end{equation}
thus for the background normalization $c$ a value has to be specified in 
\emph{all} cases.

The four parameters $a,b,c,d$ have been introduced in order to describe
systematical uncertainties, they are so called nuisance parameters.
For the analysis of the systematical errors the so called 
 pull method is 
used~\cite{Fogli:2002pt}\footnote{In fact the pull method
was employed already in~\cite{SBvNF} before \cite{Fogli:2002pt} appeared.}. 
In the  pull method $k$ systematical errors are included by introducing 
$k$ additional variables $\zeta_k$, which will  be called 
nuisance parameters in the following. 
The nuisance parameters describe the dependence of the event rates on the 
various sources of systematical errors, \eg\ an error on the total 
normalization is included by multiplying the expected number of events in 
each bin by a factor $(1+\zeta_1)$. The variation of $\zeta_1$ in the fit 
is constrained by adding a penalty $p_1$ to the $\chi^2$-function. In case 
of a  Gau\ss ian distributed systematical error this penalty is 
given by
\begin{equation}
\label{eq:penalty}
p_i=\frac{(\zeta_i-\zeta_i^0)^2}{\sigma_{\zeta_i}^2}\,,
\end{equation}
where $\zeta_i^0$ denotes the mean and $\sigma_{\zeta_i}$ the standard 
deviation of the corresponding nuisance parameter, \ie\ the amount of 
systematical uncertainty. 
The resulting $\chi^2$ is then minimized with respect to all nuisance 
parameters $\zeta_i$ and this yields $\chi^2_\mathrm{pull}$
\begin{equation}
\chi^2_\mathrm{pull}(\boldsymbol{\lambda}):=\min_{\{\zeta_i\} } \,\, \left( 
\chi^2(\boldsymbol{\lambda},
\zeta_1, \ldots, \zeta_k)+ \sum_{j=1}^{k} p_j(\zeta_j)\right)\,,
\end{equation}
where $\boldsymbol{\lambda}$ denotes the oscillation parameters 
including the matter density
$\rho$. One advantage of the pull method is that whenever the number $N$ of 
data points is much larger than $k$, it is numerically easier to compute 
$\chi^2_\mathrm{pull}$ than to invert the $N\times N$ covariance matrix. For
the experiments considered here $N$ is typically $20$ and $k\sim 4$, thus
the pull method is numerically much faster. Moreover it is more flexible and 
allows the inclusion of systematical errors also for a 
Poissonian $\chi^2$-function.
In~\cite{Fogli:2002pt} it was shown that the pull method and the covariance
based approach are equivalent for a Gau\ss ian and linear model. In general
there is a separate $(\chi^2_\mathrm{pull})^\alpha$ for each rule $\alpha$, 
\ie\ pair of signal and background spectra, with a separate set of 
nuisance parameters $\zeta_i^\alpha$. In this case $\chi^2_\mathrm{pull}$
is the sum of all individual  $(\chi^2_\mathrm{pull})^\alpha$.
In this way the dependence on the $k$ nuisance parameters has been eliminated 
from $\chi^2_\mathrm{pull}$.

The user has the possibility to choose the set $\{\zeta_i\}$ of nuisance 
parameters which are marginalized, this achieved with the error dimension 
variable\index{Error dimension}, the different possibilities are shown in
table~\ref{tab:error_dim}. If a parameter is desginated with $+$ it will
be marginalized and therefore the corresponding error needs to have a non-zero
value. In the cases labeled (number 4 and 8) ``total rates'' 
the summation over the bins is performed \emph{before} computing 
the $\chi^2$. The case labeled (number 7) ``spectrum only'' leaves the 
normalization free ($\sigma_a=\sigma_c=\infty$) and therefore counts only the 
spectral information in the date, therefore
the settings for the normalization error will be ignored. It is furthermore
possible to extend the set of error dimensions and thus the set of possible 
systematical errors to be studied, but this requires proper C programing
and recompiling \GLOBES. The interface for user defined error dimensions is
described in detail in section~\ref{sec:ud_error_dim}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{center}
\begin{table}[hbt!]
\begin{center}
\begin{tabular}[h]{|c|cccc|c|c|}
\hline
Error dimension&$a$&$b$&$c$&$d$&Tilt/Calibration&Remarks\\
\hline
\hline
0&+&+&+&+&T&\\
2&-&-&-&-&-&\\
4&+&+&+&+&T&total rates\\
7&o&-&o&-&-&spectrum only\\
8&-&-&-&-&-&total rates\\
9&+&+&+&+&C&\\
\hline
\end{tabular}
\caption[Table of error dimensions]{\label{tab:error_dim}
Values of the error dimension variable and their meaning.
 }
\end{center} 
\end{table} 
\end{center}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In each rule it is possible to constrain the energy range from which the
data are taken for the analysis\index{Energy window}. This achieved by 
setting the energy window
variable. The purpose is mainly to allow a fine tuning of the description

of an energy calibration error. The default setting is that the energy window
coincides with the range defined by the minimal and maximal reconstructed 
energy. For the error dimension using the calibration method (C) it is 
advisable to reduce the energy window by two or three bins at both the lower
and upper end. When studying the impact of the energy calibration
error in detail and it turns out not to be small, make sure that your result
does not change by reducing the energy window by 1 or 2 bins.

\subsection{Energy resolution function}
\label{sec:energy}

The origin and meaning of the energy resolution function 
$R^c(E,E')$ was already discussed in 
section~\ref{sec:channel} and  a mathematical definition
was given in equation~\ref{eq:e_res}. Here we deal with
how to tell \GLOBES\ which energy resolution function it should
use. This also naturally involves the issue how the actual convolution
of the differntial event rate spectrum with $R_\beta^{\text{IT}}(E,E')$
is performed. First we will discuss the mathematical structure of
the problem and then the various algorithms and possibilities which
\GLOBES\ offers are introduced. In order to obtain the number of events 
$n_i^c$ from one channel $c$ in one bin $i$ one has to solve this integral
%
\begin{eqnarray}
\label{eq:events_bin}
n_i^c=N/L^2\,\int_{E_i-\Delta E/2}^{E_i+\Delta E/2} dE' 
\quad \int dE \,\, \Phi^c(E)\,
P^c(E)\,
\sigma^c(E)\,
R^c(E,E')\,
\epsilon^c(E')\,.
\end{eqnarray} 
%
The integration with respect to the reconstructed energy $E'$ can be
performed independently of the oscillation parameters and we define
the bin kernel for the $i$th bin $K_i^c$
\begin{equation}
\label{eq:kernel}
K_i^c(E):=\int_{E_i-\Delta E/2}^{E_i+\Delta E/2} dE' 
\quad R^c(E,E')\,
\epsilon^c(E')\,.
\end{equation}
With this definition equation~\ref{eq:events_bin} can be rewritten as
\begin{equation}
\label{eq:simple_int}
n_i^c=N/L^2 \int dE\quad  \underbrace{\Phi^c(E)\,
P^c(E)\,
\sigma^c(E)\,
K_i^c(E)\,}_{f(E)}. 
\end{equation}

There is no basic reason why one could not evaluate this integral directly
by the usual numerical methods. However it turns out that this may be rather
slow therefore we introduce two different approximation schemes which have
a varying degree of acuracy and speed. The simplest, fastest and
least acurate is algorithm one. The key idea is to choose a set of assumptions
which allow to evaluate the integrand of equation~\ref{eq:simple_int}
at fixed values of the energy $E$ and to use the same number of points $N$ in
the calculation as there are bins. Thus equation~\ref{eq:simple_int} is reduced
to
\begin{equation}
\label{eq:algo_one}
n_i^c=N/L^2 \Delta E \sum_{j=1}^N \quad  \Phi^c(E_j)\,
P^c(E_j)\,
\sigma^c(E_j)\,
K_i^c(E_j)\,.
\end{equation}
The advantage of using equation~\ref{eq:algo_one} is that all factors
independent of the oscillation paramters have to be evaluated only at 
values of $E$ which are known in advance, \ie\ they can be put 
into a look-up table. Another advantage is that also the probability
has to be evaluated only at previously known values of the energy, which
makes it possible to compute the transition amplitudes for all channels
in one step. The key assumption is that all involved factors are piece-wise
constant, \ie\ they do not change within one bin. This assumption seems to be
very restrictive however whenever one is analyzing simulated 
data\footnote{simulated with the same algorithm} the errors
made by violating this assumption nearly cancel between the simulated 
data and the fitted data. This algorithm is the fastest which \GLOBES\ offers
and is suitable whenever there are no fast oscillations. Otherwise severe
aliasing effects may occur which usually lead to a failure of one of the 
minimizing routines. The computation of the bin kernel $K_i^c$ is performed
by \GLOBES\, to this end it requires the specification of the number of bins
and the minimum and maximum energy. Furthemore following parameterization
for $R^c(E,E')$ is used
\begin{equation}
R^c(E,E')=\frac{1}{\sigma(E)\,\sqrt{2\pi}}\,e^{-\frac{(E-E')^2}{2\sigma^2(E)}}
\end{equation} 
where the energy resolution $\sigma(E)$ is defined by
\begin{equation}
\label{eq:sigma_e}
\sigma(E)=\alpha\cdot E + \beta \cdot \sqrt{E} +\gamma\,.
\end{equation}
The parameters $\alpha, \beta$ and $\gamma$ are provided by the user.
Furthemore the user has the possibility to provide $\epsilon(E')$ by specifying
the value of  $\epsilon(E')$ at each $E_j$ since this algroithm assumes
everything else to be constant within one bin, the  $\epsilon(E')$ are
also taken as constant within one bin. In the actual implementation the sum
in equation one only extends over those $E_j$ where $K(E_j)$ above a certain
treshold, by default $10^{-7}$.

The second available algorithm is more complicated, a little slower but
much more accurate and much less susceptible to aliasing effects. The starting
point again is equation~\ref{eq:simple_int}. The idea now is to
 have a good approximation to the integral but still keeping the fixed sampling
point in $E$. To this end we rewrite the integrand of 
equation~\ref{eq:simple_int} using the sampling theorem~\cite{NRC,Rybicki}
\begin{equation}
f(E)=\sum_{k=-\infty}^{\infty}\,f(E_k) 
\times \mathrm{sinc}\left(\frac{\pi}{h}(E-E_k)\right)+e(E)
\end{equation}
with $\mathrm{sinc}(x):=\sin(x)/x$ and $E_k=E^0+h\cdot k$. 
This the so called sampling 
representation of the integrand $f(E)$ and $e(E)$ is the error term.

@@

Most advanced step: How to create a new experiment from scratch .

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: Manual.tex
%%% End:
